Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-03-31 bbbq_article.tex
Geert wrote MHC-II to B cell to T cell part, fix #202
Process feedback

For MHC-I, 9-mers were used, as this is the length most frequently presented in MHC-I and was used in our earlier
study (\cite{bianchi2017}). For MHC-II, 14-mers were used,  as these are the most frequently occurring
epitope length (\cite{bergseng2015different}).
We used a human (UniProt ID UP000005640\_9606), 
viral (SARS-CoV-2, UniProt ID UP000464024) 
and bacterial (\emph{Mycobacterium tuberculosis} (MTb), UniProt ID UP000001584) 
reference proteome. We used TMHMM (\cite{krogh2001predicting}) to predict the topology 
of the proteins within these proteomes.
To predict the affinity of an epitope to a certain MHC molecule,
we used \verb;EpitopePrediction; (\cite{bianchi2017}) for MHC-I 
and \verb;MHCnuggets; (\cite{shao2020high}) for MHC-II.
The 13 MHC-I haplotypes used in this study are the same as 
used in our previous study (\cite{bianchi2017}).
The most frequent MHC-II haplotypes used  were selected, with a phenotypic frequency of at least 14\% in
the human population (\cite{greenbaum2011functional}),
resulting in 21 MHC-II haplotypes.

@@

Additionally, we investigated if MHC-I and MHC-II are equally prone
to present TMH-derived epitopes. 
\richel{Add exact method}
See \ref{fig:rel_presentation} and \ref{fig:rel_presentation_per_haplotype}.

In our previous work, we found that the observed pattern can be
explained from the hydrophobicity of the MHC-I binding 
cleft (\cite{bianchi2017}), we validated.
Here we do the same analysis for MHC-II \richel{Explain the analysis here}.

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.95\textwidth]{bbbq_1_smart_results/fig_f_tmh_2_panel}
  \caption{
    TMH-derived peptides are predicted to be over-presented in general,
    which is for MHC-I (left column) and MHC-II (right column),
    as well as for the proteomes of SARS-CoV-2 (top row), human (middle 
    row) and \emph{M. tuberculosis} (bottom row).
    The bars show the percentage of epitopes predicted to be 
    overlapping with TMHs.
    The red lines indicate the percentage as expected by chance.
    See supplementary tables \ref{tab:tmh_binders_mhc1} and \ref{tab:tmh_binders_mhc2}
    for the exact TMH and  epitope counts.
    \richel{Add panel with hydrophobicity}
  }
  \label{fig:bbbq_1_smart_results}
\end{figure}

@@

We next wondered if the over-representation of TMH-derived peptides would also be present for MHC-II. 
Figure \ref{fig:bbbq_1_smart_results} shows the percentages of MHC-II epitopes predicted to be overlapping 
with TMHs for our human, viral and bacterial proteomes.
See supplementary table \ref{tab:tmh_binders_mhc2} 
for the TMH and epitope counts.
We predict that TMH-derived peptides are also over-presented in most MHC-II haplotypes, and this over-presentation seems even more pronounced than for MHC-I. 
\richel{
  Showing how general our finding is (as it is explained by hydrophobicity)
  is important. See \url{https://github.com/richelbilderbeek/bbbq_article/issues/192}
}
Similar to our previous results with MHC-I, we predicted that the extent to which TMH-derived
epitopes are presented by each MHC-II haplotype is similar for the human, bacterial and viral proteome.

Plotting the normalized values of presented TMH-derived epitopes (where
$1.0$ denotes that the percentage of presented TMH-derived epitopes
matches the values as expected by chance) in \ref{fig:rel_presentation}
shows that MHC-II has a higher relative over-presentation of TMH-derived
epitopes, when compared to MHC-I.
The MHC-I molecule presents epitopes of shorter length than MHC-II.
We expect this to be of no influence in the presentation level
of TMH-derived epitopes, for each of the three proteomes.
We verified this, by normalizing the percentage of TMH-derived epitopes found
to the percentage of TMH-derived epitopes expected, 
where a value of $1.0$ denotes that the percentage of TMH-derived epitopes found
matches the percentage as expected by chance.
\richel{explain only the results found. Elaborate. Draw a conclusion}
Additionally, we investigated if MHC-I and MHC-II are equally prone
to present TMH-derived epitopes. 
\richel{Add exact method}
See \ref{fig:rel_presentation} and \ref{fig:rel_presentation_per_haplotype}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \paragraph{General pattern in epitope presentation}

@@

The R package 'tmhmm' was developed to do the similar topology
predictions as our earlier study (that used 'TMHMM'), yet in an automated way.
'TMHMM' has a restrictive software license (\cite{krogh2001predicting}) and allows a user
to download a pre-compiled executable after confirmation he/she
is in academia. The R package respects this restriction
and allows the user to install and use TMHMM from within R,
as done in this study.
'tmhmm' has been submitted to and is accepted by CRAN. \frans{wat is CRAN}

@@

To determine the additional over-presentation of TMH-derived epitopes 
in MHC-II (as compared to MHC-I), we used the same normalized percentages
and compared the averages between MHC-I and MHC-II.
Figure \ref{fig:rel_presentation_per_haplotype} shows the normalized
TMH-derived epitope presentation per haplotype.
To compare the TMH-derived over-presentation per MHC class, 
we plot the mean and standard error of the normalized values,
grouped per haplotype in figure \ref{fig:rel_presentation}.

>>>>

For MHC-I, 9-mers were used, as this is the length most frequently presented in MHC-I 
and was used in our earlier study (\cite{bianchi2017}). 
For MHC-II, 14-mers were used, 
as these are the most frequently occurring epitope length (\cite{bergseng2015different}).
We used a human (UniProt ID UP000005640\_9606), 
viral (SARS-CoV-2, UniProt ID UP000464024) 
and bacterial (\emph{Mycobacterium tuberculosis} (MTb), UniProt ID UP000001584) 
reference proteome. We used TMHMM (\cite{krogh2001predicting}) to predict the topology 
of the proteins within these proteomes.
To predict the affinity of an epitope to a certain MHC molecule,
we used \verb;EpitopePrediction; (\cite{bianchi2017}) for MHC-I 
and \verb;MHCnuggets; (\cite{shao2020high}) for MHC-II.
The 13 MHC-I haplotypes used in this study are the same as 
used in our previous study (\cite{bianchi2017}).
For MHC-II, we selected haplotypes with a phenotypic frequency 
of at least 14\% in the human population (\cite{greenbaum2011functional}),
resulting in 21 MHC-II haplotypes.

@@

In our previous work, we found that the observed pattern can be
explained from the hydrophobicity of the MHC-I binding 
cleft (\cite{bianchi2017}), we validated.
Here we do the same analysis for MHC-II 
\richel{Really need to do? If yes, explain the analysis here}.

@@

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.95\textwidth]{bbbq_1_smart_results/fig_f_tmh_2_panel}
  \caption{
    TMH-derived peptides are predicted to be over-presented in general,
    which is for MHC-I (left column) and MHC-II (right column),
    as well as for the proteomes of SARS-CoV-2 (top row), human (middle 
    row) and \emph{M. tuberculosis} (bottom row).
    The bars show the percentage of epitopes predicted to be 
    overlapping with TMHs.
    The red lines indicate the percentage as expected by chance.
    See supplementary tables \ref{tab:tmh_binders_mhc1} and \ref{tab:tmh_binders_mhc2}
    for the exact TMH and  epitope counts.
    \richel{Decide in meeting: add panel with hydrophobicity}
  }
  \label{fig:bbbq_1_smart_results}
\end{figure}

@@

We next wondered if the over-representation of TMH-derived peptides would also be present for MHC-II. 
Figure \ref{fig:bbbq_1_smart_results} shows the percentages of MHC-II epitopes predicted to be overlapping 
with TMHs for our human, viral and bacterial proteomes.
See supplementary table \ref{tab:tmh_binders_mhc2} 
for the TMH and epitope counts.
We predict that TMH-derived peptides are also over-presented in most MHC-II haplotypes, and this over-presentation seems even more pronounced than for MHC-I. 
\richel{
  Decide in meeting: 
  showing how general our finding is (as it is explained by hydrophobicity)
  is important. See \url{https://github.com/richelbilderbeek/bbbq_article/issues/192}
}
Similar to our previous results with MHC-I, we predicted that the extent to which TMH-derived
epitopes are presented by each MHC-II haplotype is similar for the human, bacterial and viral proteome.

@@

% \paragraph{Conclusion}

@@

The R package 'tmhmm' was developed to do the similar topology
predictions as our earlier study (that used 'TMHMM'), yet in an automated way.
'TMHMM' has a restrictive software license (\cite{krogh2001predicting}) and allows a user
to download a pre-compiled executable after confirmation he/she
is in academia. The R package respects this restriction
and allows the user to install and use TMHMM from within R,
as done in this study.
'tmhmm' has been submitted to and is accepted 
by the Comprehensive R Archive Network (CRAN).

@@

To determine the additional over-presentation of TMH-derived epitopes 
in MHC-II (as compared to MHC-I), we normalized the data to enable
a side-by-side comparison. 
The percentage of TMH-derived epitopes presented was normalized
to the expected percentage of TMH-derived epitopes,
where $1.0$ denotes that the percentage of presented TMH-derived epitopes
matches the values as expected by chance.
The normalized values per haplotype are shown 
in figure \ref{fig:rel_presentation_per_haplotype}.
To compare the TMH-derived over-presentation per MHC class,
we grouped the normalized values per haplotype, 
and plot the mean and standard error, as shown in figure \ref{fig:rel_presentation}.
